A load balancing scheme for massively multiplayer online games \thanks{This work was supported by the National Research Council (CNPq) and by the Coordination of Improvement of Higher Education (CAPES), both brazilian research funding agencies.}

### 2009-04-10 textos/dissertation/MTAP_LaTeX/pub/tex/latex/svjour3/global/balancing.tex

Another important point is that the main criterion we use when balancing the load of MMOG servers is the bandwidth, and not processing power. Several games, such as Age of Empires \cite{ageofempires}, include simulations of virtual environments of hundreds or thousands of entities, which are performed smoothly on current personal computers. However, if this game was multiplayer and each of these entities was controlled by a different player on the Internet, it would most likely generate a traffic ammount which would hardly be supported by a domestic connection \cite{feng2007wnn}.

%TODO: explicar porque cresce linearmente
Moreover, the upload bandwidth must be taken into account, much more than download. This happens for two reasons: first, the usage of the download bandwidth of each server node grows linearly with the number of players connected to it, while the usage of the upload bandwidth may have a quadratic growth, getting quickly overwhelmed; second, domestic connections -- which are majority in a volunteer peer-to-peer system -- usually have an upload bandwidth much lower than the download one.

There are also other issues which must not be overlooked. One of them is that the server system is probably not homogenous -- considering that it is based on volunteer resources. Therefore, one can not assume that the servers have the same amount of cpu power or network bandwidth. Another problem, specific to the algorithm proposed by \cite{lee2003sdl}, is that the criterion to allocate to each server a load equivalent to 90\% of its capacity in order to avoid constant rebalancing is weak. There are no guarantees that the server system has 11.11\% more total capacity than the necessary. On the contrary, it should adapt to situations of widespread overload. Moreover, the criterion for stopping is weak because the algorithm does not stop when the whole system is loaded above 90\% of its capacity, or if the overloaded server has no neighbors capable of receiving more load.

@@

We also use the idea of mapping the virtual environment on a graph, which will then be partitioned to distribute the workload of the game between the different servers. It is necessary first to define some terms which will be used on the proposed algorithms.

@@

	\item \textbf{World weight}: the total weight of the game, $W_(total)$, depends on how the distribution will be made. It will be used as a parameter for the partitioning of the virtual environment. It is the sum of the weight of all cells. Let $\{C_1, C_2, ..., C_w\}$ be the set of all cells in which the game world is divided, we have:

@@

The criterion to choose a cell to include in the region is to make the heaviest edges on the graph $GW$ connect vertices in the same partition, reducing the edge-cut and, thus, the overhead. In each step, it is selected the cell which not only is adjacent to a cell already present in the region, but whose edge connecting them is the heaviest possible. Figure \ref{fig:ggp} illustrates the steps of growth of a region until it reaches its load share.

@@

To perform the simulation of the load balancing, a virtual environment was simulated with various avatars in it moving according to the random waypoint model \cite{bettstetter2004spr}. However, the choice of which waypoint go was not completely random. Three hotspots were defined, and there was a probability for the havatar choose one of these hotspots as its next waypoint. The purpose of this was to force an uneven distribution of avatars in the virtual environment, putting to test the algorithms of the proposed phase 2. Those who take account of the existence of hotspots form the regions on this basis, reducing the distribution overhead.

The environment consisted of a two-dimensional space, divided into 225 cells, forming a matrix of order 15. There were 750 avatars. All cells always belonged to some region, although they could be transferred between regions. There were eight servers ($S_1, S_2, ..., S_8$), each one of them associated to a region, each of which could have 0 to all 225 cells. The capacity of all servers was different, with $P(S_i) = i \times 20000$. Thus, it was possible to test whether the distribution obeyed the criterion of proportional load balancing. Each session of simulated game was 20 minutes long. The total weight of the game was purposely set in a way that it was greater than the total capacity of the system, forcing the triggering of the load balancing.

@@

We can see that \ggp\ is the one which has fewer walk migrations, precisely because their regions are contiguous. However, the fact of not trying to minimize the transfer of cells -- the \ggp\ main objective is to minimize the overhead -- the number of migrations is still the largest of all its variations, lower only than \bfa, which was also the worst algorithm on the user migration criterion. The strategy of \ggpk\, and more strongly in the \ggpf\, to maintain the maximum possible cells during rebalancing, has as a result the lowest numbers of migration of players of all the simulated algorithms.

% deviation
Another detail to be considered is the uniformity of distribution: all servers must have a usage rate as similar as possible so that the distribution is considered fair. To measure this uniformity, the standard deviation, $\sigma$, was calculated for $u(S)$ of all servers for each simulated algorithm. Figure \ref{gra:baldeviation} shows the variation of $\sigma$ in time.

@@

Even with the servers providing the largest variation in the rate of use of its resources, the \ggp\ was the one with the least overall real weight (which includes overhead: $W_{total} + Overhead$) of all proposed algorithms. Its variant \ggpf\ had the least migrations of users, in exchange for an increase in overhead by 48\%, on the average, compared to the original version of the algorithm. Which of these two is better depends on the specific game. In a real-time game , users contantly migrating between servers can introduce delay and hinder the interaction between players and the \ggpf, as shown, presented a total number of migrations 22\% less than \ggp. Furthermore, it may not be possible to use some kind of ``graceful degradation'' to reduce the quality of the game, needing to save up the servers' resources. In this case, the \ggp\ would be the best option.

@@

It was proposed here a load balancing scheme for distributed MMOGs servers, taking into account the use of upload bandwidth of the server nodes. We considered important aspects such as quadratic growth in traffic when the avatars are close to one another, and the distribution overhead when players connected to different servers are interacting. The scheme, which is divided into three phases, proposed different algorithms for phase 2 (the balancing phase), and the \ggp\ presented the lowest overhead of all, while \ggpf\ presented the fewest migrations of players between servers.

As future work, the algorithms can be further refined, if the load balancing is considered for an average of the load in a period -- in the last minutes, for example -- rather than the instantaneous value. Thus, it would avoid unnecessary rebalancing that would be caused by an oscilating weight.

Another possible future work is to create a load balancing scheme where there will be not only one server node, but a group of server nodes in each region. In this case, it can be investigated ways to balance the load of the system, considering this two-level distribution.

>>>>

Another important point is that the main criterion we use when balancing the load of MMOG servers is the bandwidth, and not processing power. Several games, such as Age of Empires \cite{ageofempires}, include simulations of virtual environments with hundreds or thousands of entities, which are performed smoothly on current personal computers. However, if this game was multiplayer and each of these entities was controlled by a different player on the Internet, it would most likely generate a traffic ammount which would hardly be supported by a domestic connection \cite{feng2007wnn}.

%TODO: explicar porque cresce linearmente
Moreover, the upload bandwidth must be taken into account, much more than download. This happens for two reasons: first, the usage of the download bandwidth of each server node grows linearly with the number of players connected to it, while the usage of the upload bandwidth may have a quadratic growth, getting quickly overwhelmed; second, domestic connections -- which are majority in volunteer peer-to-peer systems -- usually have an upload bandwidth much lower than the download one.

There are also other issues which must not be overlooked. One of them is that the server system is probably not homogenous -- considering that it is geographically distributed with, likely, different connections to the Internet. Therefore, one cannot assume that the servers have the same amount of cpu power or network bandwidth. % Another problem, specific to the algorithm proposed by \cite{lee2003sdl}, is that the criterion to allocate to each server a load equivalent to 90\% of its capacity in order to avoid constant rebalancing is weak. There are no guarantees that the server system has 11.11\% more total capacity than the necessary. On the contrary, it should adapt to situations of widespread overload. Moreover, the criterion for stopping is weak because the algorithm does not stop when the whole system is loaded above 90\% of its capacity, or if the overloaded server has no neighbors capable of receiving more load.
In the next section, we make some definitions that will be necessary to specify our load balancing scheme.

@@

We also use the idea of mapping the virtual environment on a graph. The graph will then be partitioned to distribute the workload of the game between the different servers. It is necessary first to define some terms which will be used on the proposed algorithms.

@@

	\item \textbf{World weight}: the total weight of the game, $W_{total}$, will be used as a parameter for the partitioning of the virtual environment. It is defined as the sum of the weights of all cells. Let $\{C_1, C_2, ..., C_w\}$ be the set of all cells in which the game world is divided, we have:

@@

The criterion to choose a cell to include in the region is to make the heaviest edges on the graph $GW$ connect vertices in the same partition, reducing the edge-cut and, thus, the overhead. In each step, it is selected the cell which not only is adjacent to a cell already present in the region, but whose edge connecting them is the heaviest possible. This algorithm is based on a principle similar to the one used in GGGP \cite{karypis1999fah}, a greedy algorithm for graph partitioning. The objective of GGGP is to split a graph in two partitions of the same weight, while reducing the edge-cut with the heaviest edge heuristics. Figure \ref{fig:ggp} illustrates the steps of growth of a region until it reaches its load share.

@@

To perform the simulation of the load balancing, a virtual environment was simulated with various avatars in it moving according to the random waypoint model \cite{bettstetter2004spr}. However, the choice of which waypoint to go was not completely random. Three hotspots were defined, and there was a probability for the avatar to choose one of these hotspots as its next waypoint. The purpose of this was to force an uneven distribution of avatars in the virtual environment, putting to test the load balancing algorithms. Those who take account of the existence of hotspots form the regions on this basis, reducing the distribution overhead.

As we said in section \ref{sec:related}, we also compare our algorithms to the one proposed by Ahmed and Shirmohammadi \cite{ahmed2008mol}. However, it is very important to observe that some changes have been made before simulating. First, the authors divide the virtual environment in hexagonal cells, and we simulate their algorithm on a grid-like division, with square cells. Moreover, as we consider an heterogeneous system, each server $S_i$ presents its own message rate threshold, $T_i^m$.
%
Nonetheless, we keep what we consider the core of their approach, which is the selection of the smallest cell cluster managed by the overloaded server, followed by the selection of the cell of that cluster which has the lowest interaction with other cells of the same cluster and, finally, the transference of the selected cell to the least loaded server.

The environment consisted of a two-dimensional space, divided into 225 cells, forming a matrix of order 15. There were 750 avatars. Each cell always belonged to some region, although it could be transferred to another region. There were eight servers ($S_1, S_2, ..., S_8$), each one of them associated to a region, each of which could have 0 to all 225 cells. The capacity of all servers was different, with $P(S_i) = i \times 20000$. Thus, it was possible to test whether the distribution obeyed the criterion of proportional load balancing. Each session of simulated game was 20 minutes long. The total weight of the game was purposely set in a way that it was greater than the total capacity of the system, forcing the triggering of the load balancing.

@@

We can see that \ggp\ is the one which has fewer walk migrations, precisely because their regions are contiguous. However, the fact of not trying to minimize the transfer of cells -- the \ggp\ main objective is to minimize the overhead -- the number of migrations is still the largest of all its variations, lower only than \bfa, which was also the worst algorithm on the user migration criterion. The strategy of \ggpk, and more strongly in the \ggpf, to maintain the maximum possible number of cells when rebalancing, has as a result the lowest numbers of migration of players of all the proposed algorithms.

The algorithm proposed by Ahmed had even less migrations than \ggpf. Though the number of walk migrations is the second highest -- as expected from a more fragmented partitioning --, the number of still migrations is almost negligible. We believe that a clever choice of cells to transfer led to this result. Cells which interact less with other cells will, probably, contain less avatars and, then, trasferring them will not cause many player migrations.

% deviation
Another detail to be considered is the uniformity of distribution: all servers must have an usage rate as similar as possible so that the distribution is considered fair. To measure this uniformity, the standard deviation, $\sigma$, was calculated for $u(S)$ of all servers for each simulated algorithm. Figure \ref{gra:baldeviation} shows the variation of $\sigma{}(u(S))$ over time.

@@

To summarize the evaluated algorithms, we have: \ggp, which introduced an overhead that increased the system load by 13\% on the average, but it presents the second highest number of migrations of players, and the least fair load distribution; \ggpk, whose overhead increases the system load by 28\%, but it presents the third best number of migrations and its distribution is one the most fair; \ggpf, whose load increase is of 22\%, with the second lowest number of migrations and is also one of the most fair algorithms; and we also have \bfa, which increased the load by 34\%, with the highest number of migrations and not as fair as \ggpk\ or \ggpf. Finally, Ahmed's algorithm increased the system load by 29\%, on the average, with the least number of migrations and a fairness level varying a lot over time.

Although the servers presented the largest variation in the resources usage rate, the \ggp\ was the one with the least overall real weight (which includes overhead: $W_{total} + OverHead$) of all simulated algorithms. Its variant \ggpf\ had the second lowest number of user migrations, along with the second lowest overhead and a fair load distribution, on the average, when compared to the other algorithms. Which of these two is better depends on the specific game. In a real-time game, users constantly migrating between servers can introduce delay and hinder the interaction between players.  \ggpf, or Ahmed's algorithm, would be more suited for this situation. However, when it is not possible to use some kind of ``graceful degradation'' to reduce the quality of the game and save up the servers' resources, \ggp\ would be the best option.

@@

It was proposed here a load balancing scheme for distributed MMOG servers, taking into account the use of upload bandwidth of the server nodes. We considered important aspects such as the quadratic growth of the traffic when the avatars are close to one another, and the distribution overhead when players connected to different servers are interacting. The scheme, which is divided into three phases, proposed different algorithms for phase 2 (the balancing phase), and \ggp\ presented the lowest overhead of all, while \ggpf\ presented the second fewest migrations of players between servers, along with the second lower overhead introduced and a fair load distribution. For this reason, we recommend the use of \ggpf\ for most cases. However, in some situations where the system load must be reduced as much as possible, \ggp\ would perform better.

As a future work, %the algorithms can be further refined, if the load balancing is considered for an average of the load in a period -- in the last minutes, for example -- rather than the instantaneous value. Thus, it would avoid unnecessary rebalancing that would be caused by an oscilating weight. Another possible future work is to 
one could create a load balancing scheme where there will be not only one server node, but a group of server nodes managing each region. In this case, ways to balance the load on the system, considering this two-level distribution, may be investigated.
